---
title: "COVID-19 Data Cleaning"
description: "Load, clean, and join county-level COVID-19 case data with US Census population sizes."
---

## Load Dependencies

```{r}
library(renv)
library(tidyverse)
library(readr)
```

## Load in Data

Read the raw county-wise COVID-19 case counts and the US Census population
sizes into R.

```{r}
raw_cases <- read_csv("01_data/usa_county_wise.csv")
head(raw_cases)

raw_pop <- read_csv("01_data/population_sizes.csv")
head(raw_pop)
```

## Clean Data for Joining

Standardize the `geographic_area` column in both datasets so that county names
match across the two tables. We keep only rows with a valid 5-digit FIPS code
to exclude state-level aggregates.

```{r}
cases <- raw_cases |>
  filter(nchar(FIPS) == 5) |>
  mutate(
    geographic_area = Combined_Key |>
      str_remove(", US$") |>
      str_remove(" County") |>
      str_trim()
  ) |>
  select(geographic_area, Confirmed, Deaths, Date)

pop <- raw_pop |>
  mutate(
    geographic_area = geographic_area |>
      str_remove(" County") |>
      str_trim()
  )
```

## Join Case Counts with Population Sizes

Left-join the case counts onto the population table and drop rows where no
matching population was found.

```{r}
head(cases)
head(pop)

combined <- cases |>
  left_join(pop, by = c("geographic_area" = "geographic_area")) |>
  filter(!is.na(population_size))

head(combined)
```

## Convert Date Column

Parse the `Date` column from its original `MM/DD/YY` string format into an R
`Date` object so that it can be used in time-series calculations.

```{r}
combined <- combined |>
  mutate(Date = as.Date(Date, format = "%m/%d/%y"))
```

## Calculate Incidence

Compute daily new cases (incidence) for each county as the first difference of
cumulative confirmed cases. Negative differences (data corrections) are clamped
to zero.

```{r}
combined <- combined |>
  group_by(geographic_area) |>
  arrange(Date) |>
  mutate(Incidence = Confirmed - lag(Confirmed, default = 0)) |>
  mutate(Incidence = ifelse(is.na(Incidence) | Incidence < 0, 0, Incidence)) |>
  ungroup()

top_50_incidence <- combined |>
  arrange(desc(Incidence)) |>
  slice_head(n = 50)

tail(top_50_incidence)
```

## Visualize Data

Plot daily incidence over time for counties with more than 1,000 new cases on
at least one day, using a log-10 scale on the y-axis to accommodate the wide
range of values.

```{r}
combined |>
  filter(Incidence > 1000) |>
  ggplot(aes(x = Date, y = Incidence, color = geographic_area)) +
  geom_line() +
  labs(
    title = "COVID-19 Incidence Over Time by County",
    x = "Date",
    y = "Daily Incidence"
  ) +
  scale_y_log10() +
  theme_minimal()


head(combined)
```
