---
title: "Bayesian SIR Model Fitting"
description: "Fit a Bayesian SIR model to county-level COVID-19 incidence data using CmdStanR."
---

## Load Dependencies

```{r}
library(cmdstanr)
library(bayesplot)
library(posterior)
```

## 1) Compile the Model

Compile the Stan SIR model from `model.stan`. CmdStanR will translate the Stan
code to C++ and compile it the first time; subsequent runs reuse the compiled
binary.

```{r}
mod <- cmdstan_model("03_modeling/model.stan")
```

## 2) Prepare the Data

Select a single county and build the `stan_data` list required by the model.
The SIR states are expressed as fractions of the total population, so we
compute `S0_frac` and `I0_frac` from the raw counts.

```{r}
# Select a specific county for modeling
county <- "Young, Texas"

area_data <- combined |>
  filter(geographic_area == county) |>
  arrange(Date)

# Define population and starting state
# We use the population_size from your data
pop <- area_data$population_size[1]
i0 <- 1 # Initial number infected
s0 <- pop - i0 # Susceptibles

stan_data <- list(
  T = nrow(area_data),
  t0 = 0,
  ts = 1:nrow(area_data), # timesteps
  cases = as.integer(area_data$Incidence), # observed daily new cases
  population = pop,
  S0_frac = s0 / pop,
  I0_frac = i0 / pop
)
```

## 3) Fit the Model

Run MCMC sampling with 4 chains in parallel. `adapt_delta = 0.95` reduces the
step size to improve sampling in regions of high curvature.

```{r}
fit <- mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  adapt_delta = 0.95
)
```

## 4) Check Diagnostics & Summary

Inspect posterior summaries for the key epidemiological parameters: the
transmission rate (`beta`), recovery rate (`gamma`), basic reproduction number
(`R0`), reporting rate (`rho`), and the negative-binomial overdispersion
parameter (`phi`).

```{r}
print(fit$summary(c("beta", "gamma", "R0", "rho", "phi")))
```

## 5) Visualizations

Plot MCMC trace plots for `beta`, `gamma`, and `R0` to assess chain mixing and
convergence.

```{r}
draws <- fit$draws(c("beta", "gamma", "R0"))

# Generate plots
mcmc_trace(draws)
```
